---
title: "final_models"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
final_figs_path <- file.path("analyses_manuscript/figures/final_manuscript_figures")
```

```{r}
library(sjPlot)
library(tidyverse)
library(ggpubr)
library(ggrepel)
library(effects)
library(cowplot)
library(lme4)
library(glmmTMB)
library(car)
library(emmeans)
library(stringr)
library(performance)

load("analyses_manuscript/theme_function.R")
```

```{r}
analysis_df <- read.csv("analyses_manuscript/processed_data/analysis_variables.csv") %>%
  mutate(year = as.factor(year)) %>%
  mutate(log_shannon_native = log(shannon_native + 1)) %>%
  mutate(shannon_native_backtransform = exp(log_shannon_native)-1)
```

### Function to calculate p-values

```{r}
format_pval <- function(pval) {
  
  # Assess significance of p-values
  pval_r <- ifelse(
    pval < 0.001,
    "< 0.001",
    round(pval, digits = 3)
  )
  
  significance <- ifelse(pval < 0.05,
                       "*",
                       "N.S.")
  
  pval_signif <- paste(pval_r, significance, sep = " ")
    
  return(pval_signif)
  
}
```

```{r}
test_significance <- function(fitted_model) {
  
  model_p_values <- list()
  
  # Hypothesis testing with Anova() chisq test
  chisqAnova <- car::Anova(fitted_model, component = "cond")
  
  # Extract p-value for the treatment*year interaction term
  pval_int <- chisqAnova$'Pr(>Chisq)'[3]
  pval_year <- chisqAnova$'Pr(>Chisq)'[2]
  
  # print(pval_year)
  
  pval_int_f <- format_pval(pval_int)
  model_p_values[['int_pval']] <- pval_int_f
  
  pval_year_f <- format_pval(pval_year)
  model_p_values[['year_pval']] <- pval_year_f
  
  return(model_p_values)
  
}
```

### Function to plot model

```{r}
plot_model <- function(response_variable, response_name, effects_table, dataset, fitted_model) {

  # Test significance of parameters
  test <- test_significance(fitted_model)
  
  pval_int <- test[['int_pval']]
  pval_year <- test[['year_pval']]
  
  # Format label
  significance_label = paste(
    "treatment x year p-value:", pval_int, "\n",
    "year p-value:", pval_year,
    sep = " ")
  
  # Make plot
  model_plot <- effects_table %>%
    ggplot(aes(
      x = year,
      y = fit,
      fill = treatment,
      color = treatment,
      group = treatment
    )) +
    geom_jitter(data = dataset,
                aes_string(x = "year",
                y = response_variable,
                color = "treatment"),
                alpha = 0.3,
                shape = 16,
                width = 0.3,
                height = 0
                ) +
    geom_errorbar(
      aes(ymin = lower_CL, ymax = upper_CL),
      width = 0.2,
      size = 1,
      show.legend = FALSE
    ) +
    geom_line(linewidth = 1, show.legend = TRUE) +
    geom_point(
      shape = 21,
      size = 4,
      color = "black",
      show.legend = FALSE
    ) +
    scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
    scale_color_manual(
      values = c("antiquewhite4", "darkgoldenrod2"),
      guide = guide_legend(override.aes = list(alpha = 1))
    ) +
  labs(subtitle = response_name) +
  ylab(response_name) +
  xlab("Year") +
  theme_Publication() +
  # Modify our project plotting theme for this use case 
  theme(legend.position = "right",
        plot.title = element_text(size = 12),
        plot.subtitle = element_text(face = "bold"),
        axis.title.y = element_text(face = "plain")) +
  # Add p-values from models
  # annotate("text", x = Inf, y = Inf, label = paste(significance_label), vjust = 2, hjust = 1, size = 3.5)
  labs(caption=paste(significance_label))
  
  print(model_plot)

  return(model_plot)
}

```

# Final global models

### Species richness

```{r}
sprich_fm <- glmer(sprich ~ treatment*year + (1|block/plot),
                   family = "poisson",
                   data = analysis_df)

emmeans_sprich_fm <- emmeans(sprich_fm, ~ treatment*year, component = "cond", type="response") %>%
  as.data.frame() %>%
  mutate(fit = rate,
         se = SE,
         lower_CL = asymp.LCL,
         upper_CL = asymp.UCL
         )
emmeans_sprich_fm

p_sprich <- plot_model(response_variable = "sprich",
           response_name = "Species richness",
           effects_table = emmeans_sprich_fm,
           dataset = analysis_df,
           fitted_model = sprich_fm
           )


p_sprich

```

### Native species richness

```{r}
sprich_native_fm <- glmer(sprich_native ~ treatment*year + (1|block/plot),
                   family = "poisson",
                   data = analysis_df)

emmeans_sprich_native_fm <- emmeans(sprich_native_fm, ~ treatment*year, component = "cond", type="response") %>%
  as.data.frame() %>%
  mutate(fit = rate,
         se = SE,
         lower_CL = asymp.LCL,
         upper_CL = asymp.UCL
         )
emmeans_sprich_native_fm

p_sprich_native <- plot_model(response_variable = "sprich_native", 
           response_name = "Native species richness",
           effects_table = emmeans_sprich_native_fm,
           dataset = analysis_df,
           fitted_model = sprich_native_fm)
```

### Shannon

```{r}
shannon_fm <- lmer(shannon ~ treatment*year + (1|block/plot),
                   data = analysis_df)


emmeans_shannon_fm <- emmeans(shannon_fm, ~ treatment*year, component = "cond", type="response") %>%
  as.data.frame() %>%
  mutate(fit = emmean,
         se = SE,
         lower_CL = lower.CL,
         upper_CL = upper.CL
         )
emmeans_shannon_fm

p_shannon <- plot_model(response_variable = "shannon",
           response_name = "Shannon diversity",
           effects_table = emmeans_shannon_fm,
           dataset = analysis_df,
           fitted_model = shannon_fm)
```

### Shannon Native

```{r}
shnat_fm <- glmmTMB(log_shannon_native ~ treatment*year + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family = beta_family())


emmeans_shannon_native_fm <- emmeans(shnat_fm, ~ treatment*year, component = "cond", type="response") %>%
  as.data.frame() %>%
  # Backtransform
  mutate(fit = (exp(response) -1),
         se = (exp(SE) -1),
         lower_CL = (exp(asymp.LCL) -1),
         upper_CL = (exp(asymp.UCL) -1)
         )
emmeans_shannon_native_fm

p_shannon_native <- plot_model(response_variable = "shannon_native", 
           response_name = "Native Shannon diversity",
           effects_table = emmeans_shannon_native_fm,
           dataset = analysis_df,
           fitted_model = shnat_fm)

```

### Exotic forb

```{r}
ef_fm <- glmmTMB(exotic_forb ~ treatment*year + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family= beta_family()
)

emmeans_ef_fm <- emmeans(ef_fm, ~ treatment*year, component = "cond", type="response") %>%
  as.data.frame() %>%
  mutate(fit = response,
         se = SE,
         lower_CL = asymp.LCL,
         upper_CL = asymp.UCL
         )
emmeans_ef_fm

p_exotic_forb <- plot_model(response_variable = "exotic_forb", 
           response_name = "Exotic forb abundance",
           effects_table = emmeans_ef_fm,
           dataset = analysis_df,
           fitted_model = ef_fm
           )
```

### Native forb

```{r}
nf_fm <- glmmTMB(native_forb ~ treatment*year + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family= beta_family()
)

emmeans_nf_fm <- emmeans(nf_fm, ~ treatment*year, component = "cond", type="response") %>%
  as.data.frame() %>%
  mutate(fit = response,
         se = SE,
         lower_CL = asymp.LCL,
         upper_CL = asymp.UCL
         )
emmeans_nf_fm

p_native_forb <- plot_model(response_variable = "native_forb", 
           response_name = "Native forb abundance",
           effects_table = emmeans_nf_fm,
           dataset = analysis_df,
           fitted_model = nf_fm
           )
```

## Annotate and export arranged plots.

Add results from single year models to the plots.

```{r}
single_year_model_results <- read.csv("analyses_manuscript/stats_results/single_year_model_results.csv")

single_year_model_info <- single_year_model_results %>%
  mutate(
    response_var = substr(model, 1, nchar(model) - 4),
    year = substr(model, nchar(model) - 3, nchar(model))
  ) %>%
  mutate(
    response_var_f = case_when(
      response_var == "sprich_fm_" ~ "sprich",
      response_var == "sprich_native_fm_" ~ "sprich_native",
      response_var == "shannon_fm_" ~ "shannon",
      response_var == "shnat_fm_" ~ "shannon_native",
      response_var == "ef_fm_" ~ "exotic_forb",
      response_var == "nf_fm_" ~ "native_forb"
    )
  )
single_year_model_info

  
```

Adjust axis limits for variables on the same scale.

```{r}
# Richness plots
richness_max <- max(max(analysis_df$sprich), max(analysis_df$sprich_native))

p_sprich_adj <- p_sprich + 
  scale_y_continuous(limits = c(0, richness_max)) + 
  ylab("richness")

p_sprich_native_adj <- p_sprich_native + 
  scale_y_continuous(limits = c(0, richness_max)) + 
  ylab("richness")

# Shannon plots
H_max <- max(max(analysis_df$shannon), max(analysis_df$shannon_native))

p_shannon_adj <- p_shannon + 
  scale_y_continuous(limits = c(0, H_max)) + 
  ylab("H'")

p_shannon_native_adj <- p_shannon_native + 
  scale_y_continuous(limits = c(0, H_max)) + 
  ylab("H'")

# Relative abundance plots
relabund_max <- max(max(analysis_df$native_forb), max(analysis_df$exotic_forb))

p_native_forb_adj <- p_native_forb + 
  scale_y_continuous(limits = c(0, relabund_max)) + 
  ylab("relative abundance")

p_exotic_forb_adj <- p_exotic_forb + 
  scale_y_continuous(limits = c(0, relabund_max)) + 
  ylab("relative abundance")

```

```{r}
# Make some aesthetic changes to the plots

all_plots_list <- list(
  p_sprich = p_sprich_adj,
  p_sprich_native = p_sprich_native_adj,
  p_shannon = p_shannon_adj,
  p_shannon_native = p_shannon_native_adj,
  p_exotic_forb = p_exotic_forb_adj,
  p_native_forb = p_native_forb_adj
)

all_plots_with_single_year_info_list <- list()

# Add headers
panel_headers <- c("(a)", "(b)", "(c)", "(d)", "(e)", "(f)")

# Function to remove y-labels
remove_x_labels <- function(ggplot_obj) {
    ggplot_obj + theme(axis.title.x = element_blank())
}

# Function to change y-axis label language for improved clarity
richness_to_species_per_plot <- function(ggplot_obj) {
  y_label <- ggplot_obj[["labels"]][["y"]]
  
  if (str_detect(y_label, "richness")) {
    new_y_label <- str_replace(y_label, 
                               "richness", 
                               "species per plot")
    
    ggplot_obj[["labels"]][["y"]] <- new_y_label
    
  }
  
  return(ggplot_obj)
}

# Test
richness_to_species_per_plot(p_sprich)
richness_to_species_per_plot(p_sprich_native)

```

```{r}
# Update plots

for (i in seq_along(names(all_plots_list))) {
  
  plot_name <- names(all_plots_list)[i]
  plot <- all_plots_list[[plot_name]]
  
  response_var <- substring(plot_name, first = 3)
  print(response_var)
  
  # Calculate the y-location of where the labels should be
  
  ## if you want on the bottom
  # plot_label_height <- 0
  
  ## if you want above data
  plot_y_vals <- plot[['data']][['upper_CL']]
  plot_y_vals_max <- max(plot_y_vals)
  plot_label_height <- plot_y_vals_max + (plot_y_vals_max)*0.02
  # print(paste("Plot label y-value is: ", plot_label_height))
  
  # Extract the information to annotate
  results_for_that_response_var <- single_year_model_info %>%
    filter(response_var_f == {{response_var}}) %>%
    mutate(label_height = plot_label_height)

  # Add single-year model p-values to the plot
  plot_with_single_year_significance <- plot +
    geom_text_repel(
      inherit.aes = FALSE,
      data = results_for_that_response_var,
      aes(
        x = year,
        y = label_height,
        label = paste("single-year model:", "\n", p_value_f, significance, sep = " ")
      ),
      size = 3.5,
      nudge_y = 0.01,
      segment.size = 0
    )

  # Add header alphabetical labels to the plot
  plot_with_header <- plot_with_single_year_significance +
    labs(title = panel_headers[i]) +
    # Re-apply theme to correct any style elements overwritten by the annotation code
    theme_Publication() +
    theme(plot.title = element_text(hjust = 0)) +
    theme(plot.subtitle = element_text(size = rel(1.2), face = "bold"))

  
  # Remove "year" y-label where not needed
  panel_header <- panel_headers[i]
  print(panel_header)
  
  redundant_year_labels <- c("(a)", "(b)", "(c)", "(d)")
  
  if (panel_header %in% redundant_year_labels) {
    print("yes")
    plot_with_header <- remove_x_labels(plot_with_header)
    
  }
  
  # Replace "richness" with "species per plot" for clarity, where needed
  plot_with_updated_y_label <- richness_to_species_per_plot(plot_with_header)
  
  all_plots_with_single_year_info_list[[plot_name]] <- plot_with_updated_y_label 
  
}

all_plots_with_single_year_info_list
```

```{r}
arranged_global_model_plots <- ggarrange(plotlist = all_plots_with_single_year_info_list,
                                         legend = "bottom",
                                         common.legend = TRUE,
                                        ncol = 2, nrow = 3)

arranged_global_model_plots
# Viewing this causes a view error due to the large size of all the elements
# However the export below will produce a good result

```

```{r}
ggsave(filename = "fig_3_global_models.pdf",
         plot = arranged_global_model_plots,
         path = final_figs_path,
         width = 6.5,
         height = 9,
          scale = 1.5)

ggsave(filename = "fig_3_global_models.jpg",
         plot = arranged_global_model_plots,
         path = final_figs_path,
         width = 6.5,
         height = 9,
          scale = 1.5)
```

## Table of model results

Save model results for all models.

```{r}
final_models_list <- list(sprich_fm = sprich_fm,
                     sprich_native_fm = sprich_native_fm,
                     shannon_fm = shannon_fm, 
                     shnat_fm = shnat_fm, 
                     ef_fm = ef_fm, 
                     nf_fm = nf_fm)

models_with_results_list <- list()
for (model_name in names(final_models_list)) {
  
  model_info <- list()
  
  # Extract model object
  model <- final_models_list[[model_name]]
  
  # Add model to a list of results for the model
  model_info[[model_name]] <- model

  # Add to Anova results to that list
  model_results <- car::Anova(model, component = "cond")
  model_info[["results"]] <- model_results
  
  models_with_results_list[[model_name]] <- model_info
 
}

models_with_results_list
```

```{r}
make_results_table <- function(list_of_model_results) {
  
  results_table <- tibble(
    predictor = character(),
    Chisq = numeric(),
    Df = numeric(),
    `Pr(>Chisq)` = numeric(),
    response_variable = character()
    ) %>%
    as_tibble()
  
  # Extract the model results from each model and put it in a table
  for (model in list_of_model_results) {

    results <- model[["results"]]

    response_var <- attr(results, "heading")[2]


    results_tibble <- results %>%
      as.data.frame() %>%
      rownames_to_column("predictor") %>%
      as_tibble() %>%
      mutate(response_variable = response_var) %>%
      mutate(response_variable = sub("Response: ", "", response_variable)) %>%
      mutate(model_type = "global") %>%
      mutate(p_value_pretty = format_pval(`Pr(>Chisq)`))

    results_table <- results_table %>%
      bind_rows(results_tibble)

  }
  
  return(results_table)
  
}
```

```{r}
model_results_table <- make_results_table(models_with_results_list)

model_results_table
```

Export the table.

```{r}
write.csv(
  model_results_table,
  file = "analyses_manuscript/stats_results/global_model_results.csv",
  row.names = FALSE
)
```
