---
title: " Species cover analysis"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options: 
  chunk_output_type: inline
---

```{r}
final_figs_path <- file.path("analyses_manuscript/figures/final_manuscript_figures")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# options(scipen = 999)
```

```{r}
library(tidyverse)
library(ggpubr)
library(rstatix)
library(performance)
library(lme4)
library(glmmTMB)
library(car)
library(sjPlot)
library(ggeffects)

load("analyses_manuscript/theme_function.R")

species_list <- read.csv("analyses_manuscript/raw_data/20230918_CHRRP_species_list.csv")

analysis_df <- read.csv("analyses_manuscript/processed_data/analysis_variables.csv") %>%
  mutate(year = as.factor(year))

areal_cover_df <- read.csv("analyses_manuscript/processed_data/areal_cover_variables.csv")

```

What are the most common species across all plots?

```{r}
areal_cover_df %>%
  dplyr::select(-plot, -year, -treatment) %>%
  summarise_all(mean) %>%
  gather() %>%
  arrange(desc(value))
```


Exploratory analysis for individual species.

```{r}
# Do some exploration for individual species
analysis_df %>%
  ggplot(aes(x = BRODIA_cover, color = treatment)) +
  facet_wrap("year") +
  geom_histogram(binwidth = 0.015)

analysis_df %>%
  ggplot(aes(x = BROMAD_cover, color = treatment)) +
  facet_wrap("year") +
  geom_histogram(binwidth = 0.015)

analysis_df %>%
  ggplot(aes(x = BRATOR_cover, color = treatment)) +
  facet_wrap("year") +
  geom_histogram(binwidth = 0.015)

analysis_df %>%
  ggplot(aes(x = SCHBAR_cover, color = treatment)) +
  facet_wrap("year") +
  geom_histogram(binwidth = 0.015)

analysis_df %>%
  ggplot(aes(x = HORMUR_cover, color = treatment)) +
  facet_wrap("year") +
  geom_histogram(binwidth = 0.015)

analysis_df %>%
  ggplot(aes(x = AMSMEN_cover, color = treatment)) +
  facet_wrap("year") +
  geom_histogram(binwidth = 0.015)
```

First, let's look at _Bromus diandrus_ abundance in response to raking.

Examine _B. diandrus abundance_ versus native forb abundance.

```{r}
species_analysis_df <- analysis_df %>%
  dplyr::select("plot", "block", "year", "treatment", "year_treatment",
    "BRODIA_cover", "BRATOR_cover", "AMSMEN_cover",
    "exotic_forb", "native_forb", "sprich", "shannon", "sprich_native", "shannon_native") %>%
  pivot_longer(cols = c("exotic_forb", "native_forb", "sprich", "shannon", "sprich_native", "shannon_native"),
               names_to = "composition_metric",
               values_to = "metric_value")

species_analysis_df
```


```{r}
p_brodia <- species_analysis_df %>%
  ggplot(
  aes(x = BRODIA_cover, y = metric_value,
      color = treatment, fill = treatment)
) +
facet_wrap(year ~ composition_metric, scales = "free") +
  scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
  scale_color_manual(
    values = c("antiquewhite4", "darkgoldenrod2"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  geom_smooth(method = "lm") +
  geom_point() +
  theme_bw()
p_brodia
```

```{r}
p_brator <- species_analysis_df %>%
  ggplot(
  aes(x = BRATOR_cover, y = metric_value,
      color = treatment, fill = treatment)
) +
facet_grid(composition_metric ~ year, scales = "free") +
  scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
  scale_color_manual(
    values = c("antiquewhite4", "darkgoldenrod2"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  geom_smooth(method = "lm") +
  geom_point() +
  theme_bw()
p_brator
```

```{r}
p_amsmen <- species_analysis_df %>%
  ggplot(
  aes(x = AMSMEN_cover, y = metric_value,
      color = treatment, fill = treatment)
) +
facet_grid(composition_metric ~ year, scales = "free") +
  scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
  scale_color_manual(
    values = c("antiquewhite4", "darkgoldenrod2"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  geom_smooth(method = "lm") +
  geom_point() +
  theme_bw()
p_amsmen
```

## Did raking alter _B. diandrus_ cover?

```{r}
# Do paired t-tests and then add to graph
t_test_results <- species_analysis_df %>%
  group_by(year) %>%
  rstatix::t_test(BRODIA_cover ~ treatment) %>%
  mutate(display_at = "raked") %>%
  mutate(label_height = 0.3) %>%
  mutate(p_formatted = case_when(
    p >= 0.05 ~ "N.S.",
    p > 0.001 & p < 0.05 ~ paste(as.character(round(p, digits = 3), "*")),
    TRUE ~ "p < 0.001*"
      ))

p_treatment_v_brodia <- species_analysis_df %>%
  # filter(year != "2021") %>% # we do not need to see 2021, no treatment was applied
  ggplot(
    aes(x = treatment, y = BRODIA_cover, color = treatment, fill = treatment)
  ) +
  geom_jitter(height = 0, width = 0.3, alpha = 0.3) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA,
               color = "black") +
  facet_wrap("year") +
  # Add t-test results
  geom_text(
    inherit.aes = FALSE,
    data = t_test_results,
    aes(
      x = display_at,
      y = 0.3,
      label = p_formatted
    )
  ) +
  # stat_pvalue_manual(t_test_results, label = "p") +
  scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
  scale_color_manual(
    values = c("antiquewhite4", "darkgoldenrod2"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  ylab(bquote(bold("Live " * bolditalic("B. diandrus") * " cover proportion"))) +
  xlab("Treatment") +
  # Formatting
  theme_Publication() +
  theme(legend.position = "none")
p_treatment_v_brodia

```

## Did raking alter native forb cover?

```{r}
# Do paired t-tests and then add to graph
t_test_results_nf <- analysis_df %>%
  group_by(year) %>%
  rstatix::t_test(native_forb_cover ~ treatment) %>%
  mutate(display_at = "raked") %>%
  mutate(label_height = 0.5) %>%
  mutate(p_formatted = case_when(
    p >= 0.05 ~ "N.S.",
    p > 0.001 & p < 0.05 ~ paste("p =", as.character(round(p, digits = 3)), "*"),
    TRUE ~ "p < 0.001 *"
      ))

p_treatment_v_native_forb_cover <- analysis_df %>%
  ggplot(
    aes(x = treatment, y = native_forb_cover, color = treatment, fill = treatment)
  ) +
  geom_jitter(height = 0, width = 0.3, alpha = 0.3) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA,
               color = "black") +
  facet_wrap("year") +
  # Add t-test results
  geom_text(
    inherit.aes = FALSE,
    data = t_test_results_nf,
    aes(
      x = display_at,
      y = 0.7,
      label = p_formatted
    )
  ) +
  # stat_pvalue_manual(t_test_results, label = "p") +
  scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
  scale_color_manual(
    values = c("antiquewhite4", "darkgoldenrod2"),
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  ylab("native forb areal cover") +
  # Formatting
  theme_Publication() +
  theme(legend.position = "none")
p_treatment_v_native_forb_cover
```

## Did _B. diandrus_ reduce native forb cover?

```{r}
# For this plot, we need a special color mapping

color_mapping <- c("2021_unraked" = "lightblue", 
                   "2021_raked" = "lightblue", 
                   "2022_unraked" = "darkgoldenrod2",
                   "2022_raked" = "antiquewhite4",
                   "2023_unraked" = "darkgoldenrod2",
                   "2023_raked" = "antiquewhite4")


p_lm_brodia_v_native_forb <- analysis_df %>%
  unite(year_treatment, year, treatment, sep = "_") %>%
  ggplot(aes(x = BRODIA_cover, y = native_forb_cover,
             color = year_treatment
             )) +
  scale_fill_manual(values = color_mapping) +
  scale_color_manual(
    values = color_mapping,
    guide = guide_legend(override.aes = list(alpha = 1))
  ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    aes(x = BRODIA_cover, y = native_forb_cover),
    inherit.aes = FALSE,
    method = "lm",
    color = "black",
    alpha = 0.6) +
  # Stats
  stat_cor(aes(x = BRODIA_cover, y = native_forb_cover), method = "pearson",
           inherit.aes = FALSE,
           p.accuracy = 0.001) +
  # Formatting
  theme_Publication() +
  theme(legend.position = "none") +
  ylab("native forb cover") +
  xlab("live B. diandrus cover") 
p_lm_brodia_v_native_forb
```

## Mixed model

### Global model

```{r}
brodia_native_forb_df <- species_analysis_df %>%
  filter(composition_metric == "native_forb") %>%
  pivot_wider(names_from = composition_metric,
              values_from = metric_value) %>%
  # Scale using min-max rescaling to avoid heteroscedasticity issues
  mutate(BRODIA_cover_rescaled = (BRODIA_cover - min(BRODIA_cover)) / 
           (max(BRODIA_cover) - min(BRODIA_cover))
) %>%
  mutate(BRODIA_cover_unscaled = BRODIA_cover_rescaled * (max(BRODIA_cover) - min(BRODIA_cover)) + min(BRODIA_cover)
         )
brodia_native_forb_df

```

Fit model.

```{r}
m_brodia_nf_zif <- glmmTMB(native_forb ~ BRODIA_cover_rescaled + (1|block/plot),
                       family = beta_family(),
                       ziformula = ~1,
                    data = brodia_native_forb_df)
summary(m_brodia_nf_zif)

performance::check_model(m_brodia_nf_zif)
car::Anova(m_brodia_nf_zif)

sjPlot::plot_model(m_brodia_nf_zif, type = "pred")
```

```{r}
# Extract the BRODIA data as a vector for unscaling
BRODIA_cover <- brodia_native_forb_df %>% 
  pull(BRODIA_cover)

# Predict
predicted_values <- ggpredict(m_brodia_nf_zif, terms = "BRODIA_cover_rescaled")

plot(predicted_values)
  
predicted_values_with_rescaled_x <- predicted_values %>%
  # Unscale back to raw values
  as.data.frame() %>%
    mutate(x = x * (max(BRODIA_cover) - min(BRODIA_cover)) + min(BRODIA_cover))

ggplot(predicted_values_with_rescaled_x, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.3)
```


```{r}
test_model <- function(fitted_model) {
  
  chisqAnova <- car::Anova(fitted_model)
  
  print(chisqAnova)
  
  # Extract p-value for the treatment*year interaction term
  pval <- chisqAnova$'Pr(>Chisq)'[1]
  pval.r <- ifelse(
    pval < 0.001,
    "< 0.001",
    round(pval, digits = 3)
  )

  # Assess significance of p-value
  significance <- ifelse(pval < 0.05,
                       "*",
                       "N.S.")
  pval_signif <- paste(pval.r, significance, sep = " ")
    
  print(pval_signif)
  
  return(pval_signif)
}
```


```{r}
plot_brodia_model <- function(fitted_model, response_variable, response_name, dataset) {
  
  # Calculate p-value
  pval = test_model(fitted_model)

  # Format p-value label
  significance_label <- paste("p-value:", pval, sep = " ")
  
  # Format plot title
  plot_title <- paste(
    paste(response_name, ",", sep = ""),
          significance_label, sep = " ") # if a different format is desired
  # plot_title <- significance_label
  
  # Compute predicted values
  predicted_values <- ggpredict(fitted_model, terms = "BRODIA_cover_rescaled")
  
  predicted_values_with_rescaled_x <- predicted_values %>%
  # Unscale back to raw values
    as.data.frame() %>%
    mutate(x = x * (max(BRODIA_cover) - min(BRODIA_cover)) + min(BRODIA_cover))

  # Make plot
  model_plot <- ggplot(predicted_values_with_rescaled_x, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.3) +
    geom_point(
      inherit.aes = FALSE,
      data = dataset,
      aes_string(
        x = "BRODIA_cover",
        y = response_variable,
        fill = "year_treatment",
        color = "year_treatment"
        ),
      alpha = 0.3,
      shape = 16,
      size = 3
      ) +
    scale_fill_manual(values = color_mapping) +
    scale_color_manual(
      values = color_mapping) +
  # labs(title = plot_title) +
  theme_Publication() +
  # ggtitle(plot_title) +
  theme(legend.position = "none",
        plot.title = element_text(size = 12)) +
  ylab(response_name) +
  xlab("Live B. diandrus cover") +
  # Add significance label
  annotate("text", x = Inf, y = Inf, label = paste(significance_label), vjust = 2, hjust = 1, size = 4)

  return(model_plot)
}
```

Fit for overall dataset.

```{r}
p_brodia_v_native_forb_lmm <- plot_brodia_model(
  fitted_model = m_brodia_nf_zif,
  response_variable = "native_forb",
  response_name = "native forb cover", 
  dataset = brodia_native_forb_df) +
  ylab(expression(bold("Native forb cover proportion"))) +
  xlab(expression(bold("Live " * bolditalic("B. diandrus") * " cover proportion")))

p_brodia_v_native_forb_lmm
  
```

### For treatment years only

Fit for treatment plots only, where there is more variation in native forb cover.

```{r}
brodia_native_forb_raked_only_df <- brodia_native_forb_df %>%
  filter(treatment == "raked",
         year != "2021")
  
m_brodia_nf_raked_only_zif <- glmmTMB(native_forb ~ BRODIA_cover_rescaled + (1|block/plot),
                       family = beta_family(),
                       ziformula = ~1,
                    data = brodia_native_forb_raked_only_df)

p_brodia_v_native_forb_raked_only_lmm <- plot_brodia_model(
  fitted_model = m_brodia_nf_raked_only_zif,
  response_variable = "native_forb",
  response_name = "native forb cover", 
  dataset = brodia_native_forb_raked_only_df
)
p_brodia_v_native_forb_raked_only_lmm
```

## Make Figure 2

Combine into two-panel figure.

```{r}
p_treatment_v_brodia <- p_treatment_v_brodia +
  ggtitle("(a)") +
  theme(
    plot.title = element_text(hjust = 0, size = 12)
  )

p_brodia_v_native_forb_lmm <- p_brodia_v_native_forb_lmm +
  ggtitle("(b)") +
  theme(
    plot.title = element_text(hjust = 0, size = 12)
  )

p_treatment_brodia_native_forb <- ggarrange(
  p_treatment_v_brodia,
  p_brodia_v_native_forb_lmm
)
  
p_treatment_brodia_native_forb

```


```{r}
ggsave(filename = "fig_2_live_brodia_and_native_forb.pdf",
         plot = p_treatment_brodia_native_forb,
         path = final_figs_path,
         width = 14,
         height = 6,
        scale = 0.8
       )

ggsave(filename = "fig_2_live_brodia_and_native_forb.jpg",
         plot = p_treatment_brodia_native_forb,
         path = final_figs_path,
         width = 14,
         height = 6,
        scale = 0.8,
       dpi=300
       )
```

## Summarise species cover

```{r}
species_mean_cover_years_treatments <- areal_cover_df %>%
  group_by(year, treatment) %>%
  summarize(across(is.numeric, mean))

species_mean_cover_years_treatments
```

```{r}
write.csv(
  species_mean_cover_years_treatments,
  file = "analyses_manuscript/stats_results/areal_cover_by_species.csv",
  row.names = FALSE
)
```

