---
title: "Calculating analysis variables"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
library(ggrepel)
library(ggpmisc)
library(knitr)
library(patchwork)

load("analyses_manuscript/theme_function.R")

```

## Load data

```{r}
data_sp_matrix <- read.csv("analyses_manuscript/processed_data/overall_comp_matrix.csv") %>%
  mutate(year = as.factor(year))

data_native_sp_matrix <- read.csv("analyses_manuscript/processed_data/native_comp_matrix.csv") %>%
  mutate(year = as.factor(year))

data_exotic_forb_matrix <- read.csv("analyses_manuscript/processed_data/exoticforb_comp_matrix.csv") %>%
  mutate(year = as.factor(year))
```

## Calculate diversity and relative abundance variables

```{r}
# Retrieve species list
data_sp_matrix %>% 
  colnames() %>% 
  sort()
```

Calculate abundances of exotic grasses, native forbs, and exotic forbs in each plot.

```{r}
# Import species list with provenance and group
species_list_df <- read.csv("analyses_manuscript/raw_data/20230918_CHRRP_species_list.csv") %>%
  rename('species_code' = 'hexcode')
species_list_df

data_form <- data_sp_matrix %>%
  pivot_longer(
    cols = where(is.numeric),
    names_to = "species_code",
    values_to = "abundance"
  ) %>%
  # Join the species-provenance/growth form group pairs
  left_join(species_list_df, by = "species_code") %>%
  # Remove BG and LIT which we will deal with later
  filter(species_code != "BG",
         species_code != "LIT") %>%
  # Create new variable for provenance/growth form groups
  unite(
    group, provenance, growth_form, sep = "_", remove = FALSE
  ) %>%
  # Calculate the total abundance for each group in each plot
  group_by(year, plot, treatment, group) %>%
  summarise(total_abundance = sum(abundance)) %>%
  # Create a new variable for year-treatment groups
  unite(year_treatment, year, treatment, remove = FALSE) %>%
  as.data.frame()

# Create new data frame with the abundances of each species group
data_form_wide <- data_form %>%
  pivot_wider(names_from = group,
              values_from = total_abundance,
              values_fill = 0)

head(data_form_wide)
```

Calculate alpha diversity statistics.

First create the matrices. It is critical that these are correct as any errors will disrupt downstream analyses.

```{r}
head(data_sp_matrix)
head(data_native_sp_matrix)
head(data_exotic_forb_matrix)

# Extract composition only from matrix
comp <- data_sp_matrix %>% 
  select(where(is.numeric)) %>%
  select(-BG,
         -LIT)

head(comp)

comp_native <- data_native_sp_matrix %>%
  select(where(is.numeric)) %>%
  select(-BG,
         -LIT)

head(comp_native)

comp_exotic_forb <- data_exotic_forb_matrix %>%
  select(where(is.numeric)) %>%
  select(-BG,
         -LIT)
head(comp_exotic_forb)
```

```{r}
# Calculate species richness and Shannon diversity for overall community
sprich <- specnumber(comp) # species richness
shannon <- diversity(comp, index = "shannon") # shannon diversity

data_overall_alpha <- data.frame(sprich, shannon) %>%
  # Add in plot info
  bind_cols(data_sp_matrix %>% select(!where(is.numeric)))
head(data_overall_alpha)

# Calculate species richness and Shannon diversity for only native species
sprich_native <- specnumber(comp_native) # species richness
shannon_native <- diversity(comp_native, index = "shannon")

data_native_alpha <- data.frame(sprich_native, shannon_native) %>%
  bind_cols(data_native_sp_matrix %>% select(!where(is.numeric)))
head(data_native_alpha)

# Calculate species richness and Shannon diversity for exotic forbs
sprich_exotic_forbs <- specnumber(comp_exotic_forb) # species richness
shannon_exotic_forbs <- diversity(comp_exotic_forb, index = "shannon")

data_exotic_forb_alpha <- data.frame(sprich_exotic_forbs, shannon_exotic_forbs) %>%
  bind_cols(data_exotic_forb_matrix %>% select(!where(is.numeric)))
head(data_exotic_forb_alpha)
```

Combine to get a final df with all of the response variables for each plot.

```{r}
data_diversity <- data_form_wide %>%
  # Join overall alpha diversity variables
  left_join(data_overall_alpha, by = c("plot", "year", "treatment", "year_treatment")) %>%
  # Join native alpha diversity variables
  left_join(data_native_alpha, by = c("plot", "year", "treatment", "year_treatment")) %>%
  # Join exotic forb alpha diversity variables
  left_join(data_exotic_forb_alpha, by = c("plot", "year", "treatment", "year_treatment")) %>%
  # Create new variables for row, column, and position of each plot
  separate(
    plot,
    into = c("row", "column", "position"),
    sep = "_",
    remove = FALSE
  ) %>%
  # Set column to numeric to ensure correct rearranging
  mutate(column = as.numeric(column)) %>%
  # Reorder by year, row, column, position
  arrange(year, row, column, position)

head(data_diversity)
```

We now have a data frame with all of our diversity variables calculated.

## Calculate litter

Import soil cover df.

```{r}
soil_cover <- read.csv("analyses_manuscript/processed_data/overall_comp_matrix.csv") %>%
  mutate(year = as.factor(year)) %>%
  select(plot, year, treatment, BG, LIT)
soil_cover
```

Join litter/soil cover with response variables.

```{r}
data_diversity_lit <- soil_cover %>%
  full_join(data_diversity, by = c("plot", "year", 'treatment'))
head(data_diversity_lit)
```

```{r}
data_diversity_lit %>%
  ggplot(aes(x=LIT)) +
  facet_wrap(treatment ~ year) +
  geom_histogram()
```

## Calculate areal cover

```{r}
areal_cover <- read.csv("analyses_manuscript/processed_data/areal_cover_variables.csv") %>%
  mutate(year = as.factor(year))
head(areal_cover)
```

```{r}
data_diversity_lit_cover <- data_diversity_lit %>%
  full_join(areal_cover, by = c('year', 'plot', 'treatment'))
head(data_diversity_lit_cover)
```

## Combine all response variables

```{r}
# Add some more relevant variables for analysis

all_analysis_variables <- data_diversity_lit_cover %>%
  # Add blocks
  unite(block, row, column, sep = "_", remove = FALSE) %>%
  # Add number of years raked
  mutate(
    years_raked = case_match(
      year_treatment,
      '2021_unraked' ~ 0,
      '2021_raked' ~ 0,
      '2022_unraked' ~ 0,
      '2022_raked' ~ 1,
      '2023_unraked' ~ 0,
      '2023_raked' ~ 2
    )
  )

head(all_analysis_variables)
```

## Basic exploratory analysis

Look at the range of values for the response variables.

```{r}
summary(all_analysis_variables)
```

### Visualize data

Create a table of summary statistics.

```{r}
all_analysis_variables_long <- all_analysis_variables %>%
  select(
    plot, year, treatment, 
         # Soil cover
         BG,
         LIT,
         # Relative abundances of groups
         exotic_forb,
         exotic_grass,
         native_forb,
         # Cover of groups
         exotic_forb_cover,
         exotic_grass_cover,
         native_forb_cover,
         # Alpha diversity
         sprich,
         sprich_native,
         shannon,
         shannon_native,
        sprich_exotic_forbs,
        shannon_exotic_forbs
    ) %>%
  pivot_longer(cols = 4:length(.),
               names_to = "metric",
               values_to = "value") %>%
  # Create formatted variable names
  mutate(metric_name_f = case_when(
    metric == 'BG' ~ 'bare ground',
    metric == 'LIT' ~ 'litter soil cover',
    
    metric == 'exotic_forb' ~ 'exotic forb relative abundance',
    metric == 'exotic_forb_cover' ~ 'exotic forb areal cover',
    
    metric == 'exotic_grass' ~ 'exotic grass relative abundance',
    metric == 'exotic_grass_cover' ~ 'exotic grass areal cover',
    
    metric == 'native_forb' ~ 'native forb relative abundance',
    metric == 'native_forb_cover' ~ 'native forb areal cover',
    
    metric == 'sprich' ~ 'species richness',
    metric == 'sprich_native' ~ 'native species richness',
    metric == 'sprich_exotic_forbs' ~ 'exotic forb species richness',
    
    metric == 'shannon' ~ 'Shannon diversity',
    metric == 'shannon_native' ~ 'native Shannon diversity',
    metric == 'shannon_exotic_forbs' ~ 'exotic forb Shannon diversity'
  ))

summary_all_variables <- all_analysis_variables_long %>%
  group_by(year, treatment, metric, metric_name_f) %>%
  summarise(
    mean = mean(value),
    sd = sd(value),
    se = sd(value) / sqrt(length(value)),
    n = length(value)
  ) %>%
  ungroup() %>%
  arrange(year, treatment, metric) %>%
  as.data.frame() %>%
  # Add column for the y-axis labels
  mutate(
    y_axis_label = case_when(
      metric == 'BG' ~ 'areal cover proportion',
      metric == 'LIT' ~ 'areal cover proportion',
      
      metric == 'exotic_forb' ~ 'relative abundance',
      metric == 'exotic_forb_cover' ~ 'areal cover proportion',
      
      metric == 'exotic_grass' ~ 'relative abundance',
      metric == 'exotic_grass_cover' ~ 'areal cover proportion',
      
      metric == 'native_forb' ~ 'relative abundance',
      metric == 'native_forb_cover' ~ 'areal cover proportion',
      
      metric == 'sprich' ~ 'species per plot',
      metric == 'sprich_native' ~ 'species per plot',
      metric == 'sprich_exotic_forbs' ~ 'species per plot',
      
      metric == 'shannon' ~ "H'",
      metric == 'shannon_native' ~ "H'",
      metric == 'shannon_exotic_forbs' ~ "H'"
    )
  )
  
summary_all_variables

```

```{r}
# Visualize all means, SEs, and raw data
p_means_all_vars <- summary_all_variables %>%
  ggplot(aes(
    x = year,
    y = mean,
    color = treatment,
    fill = treatment,
    group = treatment
  )) +
  # Add in raw data
  geom_jitter(
    inherit.aes = FALSE,
    data = all_analysis_variables_long,
    aes(
      x = year,
      y = value,
      color = treatment,
      fill = treatment,
      group = treatment
    ),
    height = 0,
    width = 0.2,
    alpha = 0.2
  ) +
  # Display means
  geom_errorbar(
    aes(ymin = mean - se,
        ymax = mean + se),
    width = 0.3,
    alpha = 0.9,
    linewidth = 0.8
  ) +
  geom_line() +
  geom_point(shape = 21, color = "black") +
  facet_wrap(
    "metric_name_f",
    nrow = 5,
    ncol = 3,
    scales = "free_y"
  ) +
  ylab("value") + # placeholder, will adjust below
  scale_color_manual(values = c(
    "raked" = "antiquewhite4",
    "unraked" = "darkgoldenrod2"
  )) +
  scale_fill_manual(values = c(
    "raked" = "antiquewhite4",
    "unraked" = "darkgoldenrod2"
  )) +
  theme_Publication() +
  theme(legend.position = "bottom") +
  labs(title = "(a) Means +/- S.E. for all diversity & group abundance variables")
  
p_means_all_vars
```

Plot again but this time with a loop to allow y-axis labels to vary.

```{r}
# This must be in a logical order for combining the plots below

all_vars <- c(
  "bare ground",
  "litter soil cover",
  "exotic forb relative abundance", 
  "exotic forb areal cover",
  "exotic grass relative abundance",
  "exotic grass areal cover",
  "native forb relative abundance", 
  "native forb areal cover",         
  "Shannon diversity",      
  "exotic forb Shannon diversity",  
  "native Shannon diversity",    
  "species richness",
  "exotic forb species richness",
  "native species richness"
)

all_vars

plot_all_vars <- function(vars, spread_metric) {
  
  plots_list <- list()
  
  for (var in all_vars) {
    
    # print(var)
    
    raw_points <- all_analysis_variables_long %>%
      filter(metric_name_f == var)
    
    summary_points <- summary_all_variables %>%
      filter(metric_name_f == var)
    
    y_label <- summary_points %>% 
      select(y_axis_label) %>%
      distinct() %>% 
      pull(y_axis_label)
    
    plot_title <- summary_points %>%
      select(metric_name_f) %>%
      distinct() %>% 
      pull(metric_name_f)
    
    # If the variable name is long, add a newline character
    if (nchar(plot_title) > 15) {
      # Split the title and insert newline after the third word
      plot_title_fmd <- if (str_count(plot_title, "\\S+") > 3) {
        str_replace(plot_title, str_c(word(plot_title, 2), " "), str_c(word(plot_title, 2), "\n"))
      } else {
        plot_title_fmd <- plot_title  # If fewer than 3 words, keep it as is
      }
    } else {
      plot_title_fmd <- plot_title
    }
    
    # Make plot  
    plot <- summary_points %>%
      ggplot(aes(
        x = year,
        y = mean,
        color = treatment,
        fill = treatment,
        group = treatment
      )) +
      # Add in raw data
      geom_jitter(
        inherit.aes = FALSE,
        data = raw_points,
        aes(
          x = year,
          y = value,
          color = treatment,
          fill = treatment,
          group = treatment
        ),
        height = 0,
        width = 0.2,
        alpha = 0.2
      ) +
      # Display means
      geom_errorbar(
        aes(ymin = mean - .data[[spread_metric]],
            ymax = mean + .data[[spread_metric]]),
        width = 0.3,
        alpha = 0.9,
        linewidth = 0.8
      ) +
      geom_line() +
      geom_point(shape = 21, color = "black") +
      # Formatting
      ylab(y_label) +
      labs(title = plot_title_fmd) +
      scale_color_manual(values = c(
        "raked" = "antiquewhite4",
        "unraked" = "darkgoldenrod2"
      )) +
      scale_fill_manual(values = c(
        "raked" = "antiquewhite4",
        "unraked" = "darkgoldenrod2"
      )) +
      theme_Publication() +
      theme(legend.position = "bottom") +
      theme(
      text = element_text(size = 9),         
      axis.title = element_text(size = 9),
      axis.text = element_text(size = 9), 
      plot.title = element_text(size = 9),
      plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm"))
    
    # print(plot)
    
    plots_list[[var]] <- plot
    
  }

  combined_plots <- ggpubr::ggarrange(plotlist = plots_list,
                                 common.legend = TRUE,
                                 legend = "right",
                                 nrow = 5,
                                 ncol = 3)
  
  return(combined_plots)

}
```

```{r}
# Make mean +/- SE plot
mean_se_plots <- plot_all_vars(vars = all_vars, spread_metric = "se")

# Add title
final_mean_se_plots <- ggpubr::annotate_figure(
    mean_se_plots,
    top = text_grob("(a) Means +/- S.E. for all diversity & group abundance variables", 
                    size = 14, 
                    face = "bold")
  )

final_mean_se_plots
```

```{r}
# Visualize standard deviation
p_sd_all_vars <- summary_all_variables %>%
  ggplot(aes(
    x = year,
    y = mean,
    color = treatment,
    fill = treatment,
    group = treatment
  )) +
  # Add in raw data
  geom_jitter(
    inherit.aes = FALSE,
    data = all_analysis_variables_long,
    aes(
      x = year,
      y = value,
      color = treatment,
      fill = treatment,
      group = treatment
    ),
    height = 0,
    width = 0.2,
    alpha = 0.2
  ) + 
  # Display means
  geom_errorbar(
    aes(ymin = mean - sd,
        ymax = mean + sd),
    width = 0.3,
    alpha = 0.9,
    linewidth = 0.8
  ) +
  geom_line() +
  geom_point(shape = 21, color = "black") +
  facet_wrap("metric_name_f",
             nrow=5,
             ncol=3,
           scales = "free_y") +
  ylab("value") +
  scale_color_manual(values = c("raked" = "antiquewhite4", "unraked" = "darkgoldenrod2")) +
  scale_fill_manual(values = c("raked" = "antiquewhite4", "unraked" = "darkgoldenrod2")) +
  theme_Publication() +
  theme(legend.position = "bottom") +
  labs(title = "(b) Means +/- S.D. for all diversity & group abundance variables")
p_sd_all_vars
```

```{r}
# Use the previous plotting function to create a figure with y-axis labels that correspond to the scale of each variable, but plotting standard deviation as the error bar

mean_sd_plots <- plot_all_vars(vars = all_vars, spread_metric = "sd")

# Add title
final_mean_sd_plots <- ggpubr::annotate_figure(
   mean_sd_plots,
    top = text_grob("(a) Means +/- S.D. for all diversity & group abundance variables", 
                    size = 14, 
                    face = "bold")
  )

final_mean_sd_plots
```

## Export S.D. & S.E. figures

```{r}
# Save Figure S3a
ggsave(
  "analyses_manuscript/figures/supplemental_figures/fig_S3a_means_se_all_response_variables.jpg",
  plot = final_mean_se_plots,
  bg = "white",
  device = "png",
  height = 10,
  width = 8.4,
  units = "in",
  scale = 1,
  dpi = 300
)

ggsave(
  "analyses_manuscript/figures/supplemental_figures/fig_S3a_means_se_all_response_variables.pdf",
  plot = final_mean_se_plots,
  bg = "white",
  device = "pdf",
  height = 10,
  width = 8.4,
  units = "in",
  scale = 1
)

# Save Figure S3b
ggsave(
  "analyses_manuscript/figures/supplemental_figures/fig_S3b_means_sd_all_response_variables.jpg",
  plot = final_mean_sd_plots,
  bg = "white",
  device = "png",
  height = 10,
  width = 8.4,
  units = "in",
  scale = 1,
  dpi = 300
)

ggsave(
  "analyses_manuscript/figures/supplemental_figures/fig_S3b_means_sd_all_response_variables.pdf",
  plot = final_mean_sd_plots,
  bg = "white",
  device = "pdf",
  height = 10,
  width = 8.4,
  units = "in",
  scale = 1
)
```

## Export data summary

Export the data frame.

```{r}
write.csv(all_analysis_variables,
          file = "analyses_manuscript/processed_data/analysis_variables.csv",
          row.names =  FALSE)
```

Export the summary statistics.

```{r}
write.csv(summary_all_variables,
          file = "analyses_manuscript/stats_results/summary_of_response_variables.csv",
          row.names =  FALSE)
```

## Basic summary numbers

### Means & Effect Sizes

How much higher was native forb abundance and exotic forb abundance in treatment plots compared to control plots across 2022 and 2023?

```{r}
raked_v_unraked_forbs <- all_analysis_variables %>%
  filter(year != 2021) %>%
  group_by(treatment) %>%
  summarise(
    native_forb_mean = mean(native_forb),
    exotic_forb_mean = mean(exotic_forb)
  ) %>%
  pivot_longer(
    cols = c(native_forb_mean, exotic_forb_mean),
    names_to = "metric"
  ) %>%
  pivot_wider(
    names_from = treatment
  ) %>%
  mutate(
    diff_raked_unraked = raked - unraked
  )

raked_v_unraked_forbs
```

What was exotic forb richness and Shannon diversity on average?

```{r}
summary_all_variables %>%
  filter(metric == "sprich_exotic_forbs") %>%
  arrange(mean)

summary_all_variables %>%
  filter(metric == "shannon_exotic_forbs") %>%
  arrange(mean)
  
```

```{r}
all_analysis_variables %>%
  pull(sprich_exotic_forbs) %>%
  mean()
```

What was the greatest difference in species richness among treatments, within years?

```{r}
sprich_among_trt_by_year <- summary_all_variables %>%
  filter(metric == "sprich") %>%
  select(year, treatment, metric, mean) %>%
  pivot_wider(
    names_from = metric,
    names_prefix = "mean_",
    values_from = mean
  ) %>%
  pivot_wider(
    id_cols = year,
    names_from = treatment,
    values_from = mean_sprich
  ) %>%
  mutate(
    diff_in_mean_sprich = raked-unraked
  )

sprich_among_trt_by_year
  
```

### Species List

```{r}
species_list_for_project <- data_sp_matrix %>%
  select(-plot, -year, -treatment, -year_treatment, # these are plot-level identifiers
         -BG, -LIT # not species
         ) %>%
  colnames() %>%
  as.data.frame() %>%
  set_names("species_code") %>%
  mutate(
    known_in_site = "found in 2021-2023 study"
  ) %>%
  full_join(species_list_df, by = "species_code") %>%
  mutate(
    known_in_site = case_when(
      is.na(known_in_site) ~ "known to be at site, found in previous study (unpublished data)",
      TRUE ~ known_in_site
    )
  ) %>%
  select(species_code, species.name, provenance, growth_form, known_in_site)
  

species_list_for_project
```

```{r}
# Cross-reference with the more raw data
long_data <- read.csv("analyses_manuscript/processed_data/long_data.csv")

species_from_long_data <- long_data %>% 
  distinct(species_code) %>%
  arrange(species_code) %>% 
  filter(species_code != "BG",
         species_code != "LIT") %>%
  mutate(check = "in raw data")

species_list_for_project %>%
  filter(known_in_site == "found in 2021-2023 study") %>%
  full_join(species_from_long_data, by = "species_code")
```

```{r}
# Export species list

write.csv(species_list_for_project,
          file = "analyses_manuscript/stats_results/species_list_for_project.csv",
          row.names =  FALSE)

```

