---
title: "Litter cover analysis"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and data

```{r}
library(tidyverse)
library(rstatix) # pipe-friendly
library(ggnewscale)
library(ggpubr)
library(performance)
library(lme4)
library(sjPlot)
library(DHARMa)

options(scipen = 5)
```

```{r}
load("analyses_manuscript/theme_function.R")

# Color map
my_color_map <- c( 
                   "unraked" = "darkgoldenrod2",
                   "raked" = "antiquewhite4",
                   "2021_unraked" = "lightblue", 
                   "2021_raked" = "lightblue", 
                   "2022_unraked" = "darkgoldenrod2",
                   "2022_raked" = "antiquewhite4",
                   "2023_unraked" = "darkgoldenrod2",
                   "2023_raked" = "antiquewhite4")

```


```{r}
analysis_df <- read.csv("analyses_manuscript/processed_data/analysis_variables.csv") %>%
  mutate(year = as.factor(year)) %>%
  mutate(log_shannon_native = log(shannon_native + 1))
```

Function for model checking.

```{r}
check_my_model <- function(fitted_model, dataset) {
  
  # Residuals versus fitted
  
    tryCatch({
      print(plot(fitted_model))
    }, error = function(e) {
      cat("Error occurred while plotting residuals versus fitted:", e$message, "\n")
  })
  
   # Residual normality
  print(plot(simulateResiduals(fittedModel = fitted_model)))
  hist(resid(fitted_model))
  
  # Residuals versus covariates
  
  boxplot(resid(fitted_model) ~ treatment, data = dataset, ylab = "Normalized residuals")
  abline(h = 0, lty = 2)
  
  boxplot(resid(fitted_model) ~ year, data = dataset, ylab = "Normalized residuals")
  abline(h = 0, lty = 2)
  
  boxplot(resid(fitted_model) ~ block, data = dataset, ylab = "Normalized residuals")
  abline(h = 0, lty = 2)
  
}
```

## Visualize first

Density plots.

```{r}
analysis_df %>%
  ggplot(aes(x=LIT, color = treatment, fill = treatment, groups = year)) +
  #facet_grid("year", scales = "fixed") +
  geom_density(alpha = 0.6) +
  #geom_dotplot() +
  scale_color_manual(values = my_color_map) +
  scale_fill_manual(values = my_color_map) +
  theme_linedraw()
```

```{r}
analysis_df %>%
  ggplot(aes(x=LIT, color = treatment, fill = treatment)) +
  facet_grid("year", scales = "fixed") +
  geom_density(alpha = 0.6) +
  # geom_dotplot(dotsize = 0.3) +
  scale_color_manual(values = my_color_map) +
  scale_fill_manual(values = my_color_map) +
  theme_linedraw()
```

Histogram.

```{r}
# Make a histogram to visualize first
analysis_df %>%
  ggplot(aes(x = LIT, fill = treatment)) +
  geom_histogram(binwidth = 0.05) +
  facet_grid(year ~ treatment) +
  scale_color_manual(values = my_color_map) +
  scale_fill_manual(values = my_color_map)

## the data is clearly not normal (it is skewed). Let's investigate further.
```

Explore the LIT variable

```{r}
qqnorm(analysis_df$LIT)
qqline(analysis_df$LIT)
```


```{r}
# Test for normality using Shapiro-Wilk Test for each group
analysis_df %>%
  group_by(year_treatment) %>%
  shapiro_test(LIT)
## p < 0.001 means the data is NOT normal. The results of the test match the histograms. Only 2022 raked is normal.
```


```{r}
# Log transform LIT cover to make it more normal
analysis_df2 <- analysis_df %>%
  dplyr::select(-BG) %>% # Since BG is 1-LIT, we don't need it anymore for the analysis)
  mutate(log_LIT = log10(LIT+0.001)) # we need to add a tiny value because there are zeros and mathematically, the log10 of zero is -infinity
head(analysis_df2)

# Visualize the new log-transformed data
analysis_df2 %>%
  ggplot(aes(x = log_LIT, fill = treatment)) +
  geom_histogram(binwidth = 0.05) +
  facet_grid(year ~ treatment) +
  scale_color_manual(values = my_color_map) +
  scale_fill_manual(values = my_color_map)

```

```{r}
# Make QQ plot to visualize normality for the log-transformed data
qqnorm(analysis_df$LIT) # the original one
qqline(analysis_df$LIT)

qqnorm(analysis_df2$log_LIT)
qqline(analysis_df2$log_LIT) # slightly better!
```

Boxplot for litter cover.

```{r}
## Make litter cover boxplot

p_lit <- analysis_df %>%
  ggplot(
    aes(x = year, y = LIT, fill = treatment, color = treatment)) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2.5) +
  scale_color_manual(values = my_color_map) +
  theme_Publication() +
  new_scale_color() +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = my_color_map) +
  facet_wrap("treatment") +
  theme(legend.position = "none") +
  xlab("year") +
  ylab("litter cover proportion")
p_lit

```

```{r}
supplementary_figs_path <- file.path("analyses_manuscript/figures/supplemental_figures")

ggsave(
  filename = "fig_S2_litter_cover_by_treatment.pdf",
  path = supplementary_figs_path,
  plot = p_lit,
  device = "pdf",
  height = 6,
  width = 10,
  units = "in",
  scale = 0.6,
  dpi = 800
)

ggsave(
  filename = "fig_S2_litter_cover_by_treatment.jpg",
  path = supplementary_figs_path,
  plot = p_lit,
  height = 6,
  width = 10,
  units = "in",
  scale = 0.6,
)
```

## Statistical tests

### Basic tests

T-tests.

```{r}
# Were historically raked plots different in litter cover in 2021?
analysis_df %>%
  filter(year == "2021") %>%
  t_test(LIT ~ treatment)
```

```{r}
# Did raking reduce litter abundance among treatments in 2022?
analysis_df %>%
  filter(year == "2022") %>%
  t_test(LIT ~ treatment)
```

```{r}
# Did raking reduce litter abundance among treatments in 2023?
analysis_df %>%
  filter(year == "2023") %>%
  t_test(LIT ~ treatment)
```

ANOVA.

```{r}
anova.lit <- aov(LIT ~ year_treatment, analysis_df)
summary(anova.lit)
TukeyHSD(anova.lit)
```

```{r}
check_model(anova.lit)
```

Looks ok.

### Mixed model

```{r}
litter.lmm <- lmer(LIT ~ treatment*year + (1|block/plot),
                   data = analysis_df)

summary(litter.lmm)

check_model(litter.lmm)

```

Model check looks ok.


```{r}
check_my_model(litter.lmm, dataset = analysis_df)
```


```{r}
Anova(litter.lmm)

plot_model(litter.lmm, type = "est", show.values = T)
```

How much did raking reduce litter cover?

```{r}
summary_litter <- analysis_df %>%
  group_by(year, treatment) %>%
  summarise(
  mean_litter = mean(LIT)
) %>%
  pivot_wider(
    names_from = treatment,
    values_from = mean_litter
  ) %>%
  mutate(
    diff_between_treatment_control = unraked - raked
  )
summary_litter
```

How much did raking reduce litter in raked plots?

```{r}
print(summary_litter %>%
  filter(year != '2021') %>%
  pull(raked) %>%
  mean())
```

```{r}
write.csv(
  summary_litter,
  file = "analyses_manuscript/stats_results/mean_litter_reduction.csv",
  row.names = FALSE
)
```

## Litter versus diversity variables

```{r}
vars <- c("sprich", "sprich_native", "shannon", "shannon_native", "native_forb", "exotic_forb", "exotic_grass")

for (var in vars) {

  regression <- ggscatter(
    analysis_df, x = "LIT", y = var,
    color = "treatment",
    add = "reg.line"
    ) +
    facet_grid(year ~ treatment) +
    scale_color_manual(values = my_color_map) +
    stat_cor() +
    theme_bw()
  
  print(regression)
  
}
```

Litter versus diversity variables across all years and treatments.

```{r}

litter_plots_list <- list()
for (var in vars) {

  regression <- ggplot(
    analysis_df, 
    aes_string(x = "LIT", y = var)
    ) +
    geom_point(size = 0.5, alpha = 0.3) +
    geom_smooth(stat = "smooth", method = "lm") +
    scale_color_manual(values = my_color_map) +
    stat_cor() +
    theme_bw()
  
  litter_plots_list[[var]] <- regression
  
}

ggarrange(plotlist = litter_plots_list)
```


