---
title: "Model selection"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and data

```{r}
library(tidyverse)
library(performance)
library(lme4)
library(car)
library(DHARMa)

library(fitdistrplus)
library(glmmTMB)
library(GLMMadaptive)

library(sjPlot)
library(emmeans)
library(ggpubr)

# Set ggplot theme
theme_set(theme_bw())
```

```{r}
analysis_df <- read.csv("analyses_manuscript/processed_data/analysis_variables.csv") %>%
  mutate(year = as.factor(year)) %>%
  mutate(year_num = case_when(
                    year == '2021' ~ 0,
                    year == '2022' ~ 1,
                    year == '2023' ~ 2)) %>%
  # filter(year != "2021") %>%
  mutate(log_shannon_native = log(shannon_native + 1))

head(analysis_df)

```


## Functions for generic model workflow

Model to test.

```{r}
m_test <- lmer(shannon ~ treatment*year + (1|block/plot),
               analysis_df)
```


### Check model

```{r}
check_my_model <- function(fitted_model, dataset) {
  
  # Residuals versus fitted
  
    tryCatch({
      print(plot(fitted_model))
    }, error = function(e) {
      cat("Error occurred while plotting residuals versus fitted:", e$message, "\n")
  })
  
   # Residual normality
  print(plot(simulateResiduals(fittedModel = fitted_model)))
  hist(resid(fitted_model))
  
  # Residuals versus covariates
  
  boxplot(resid(fitted_model) ~ treatment, data = dataset, ylab = "Normalized residuals")
  abline(h = 0, lty = 2)
  
  boxplot(resid(fitted_model) ~ year, data = dataset, ylab = "Normalized residuals")
  abline(h = 0, lty = 2)
  
  boxplot(resid(fitted_model) ~ block, data = dataset, ylab = "Normalized residuals")
  abline(h = 0, lty = 2)
  
}
```

```{r}
check_my_model(m_test, dataset = analysis_df)
```

### Test model

```{r}
test_my_model <- function(fitted_model) {
  
  model_outputs <- list()
  
  # Hypothesis testing with Anova() chisq test
  chisqAnova <- Anova(fitted_model)
  
  model_outputs[["chisquare_anova_test"]] <- chisqAnova

  # Extract p-value for the treatment*year interaction term
  pval <- chisqAnova$'Pr(>Chisq)'[3]
  pval.r <- signif(pval, digits = 3)
  
  model_outputs[["p-value"]] <- pval.r

  # Assess significance of p-value
  significance <- ifelse(pval < 0.05,
                       "*",
                       "N.S.")
  
  model_outputs[["significance"]] <- significance
  
  print(model_outputs)
  
  return(model_outputs)
  
}
```

Test function.

```{r}
test_my_model(m_test)
```

### Simulate from model

```{r}
simulate_my_model <- function(model, rv) {

  all_sims <- list()
  
  for (i in c(1:100)) {
    
    sim <- simulate(model, nsim=1)
    
    all_sims[[i]] <- sim
    
  }
  
  # Combine the vectors into a data frame
  sim_data <- data.frame(
    value = unlist(all_sims),
    sim_num = rep(1:100, each = 239)
  )
  
  sim_data$sim_num_name <- paste("simulation_", sim_data$sim_num, sep = "")
  sim_data$sim_num_name <- as.factor(sim_data$sim_num_name)

  print(head(sim_data))
  
  # Plot simulated versus actual
  sim_plot <- ggplot() +
    geom_histogram(data = sim_data, aes(x = value, fill = sim_num_name),
      alpha = 0.6,
                   position = "identity",
                   color = "black") +
    scale_fill_grey() +
    # Plot real data on top
    geom_histogram(data = analysis_df, aes(x = .data[[rv]]), fill = "red", alpha = 0.5) +
    labs(title = "Simulations using model (greys) versus actual data (red)", x = "response variable", y = "frequency") +
    theme(legend.position = "none")
  
  print(sim_plot)

  return(sim_plot)

}

```

Test function.

```{r}
# simulate_my_model(m_test, rv = "shannon")
```


```{r}
fit_and_check_glmm_model <- function(response_variable, model_structure, dataset) {
  
  # Create list to store all outputs for one model
  model_outputs <<- list()
  
  model_outputs[["response_variable"]] <<- response_variable
  
  # Create the formula for the linear mixed effects model
  formula <- as.formula(paste(response_variable, model_structure))
  print(formula)
  
  model_outputs[["formula"]] <<- formula

  # Fit the linear mixed effects model
  fitted_model <- glmer(formula, data = dataset,
                        family = "poisson")
  
  model_outputs[["fitted_model"]] <<- fitted_model
  
  print(summary(fitted_model))
    
  # Model diagnostics
  check_my_model(fitted_model = fitted_model, dataset = dataset)
  
  # Checks unique to GLMMs
  print(check_zeroinflation(fitted_model))
  print(check_overdispersion(fitted_model))
    
  # Other assumptions
  print(check_model(fitted_model))
  
  # Test
  test_my_model(fitted_model = fitted_model)
  
  # Simulate
  simulate_my_model(fitted_model, rv = response_variable)
  
  return(model_outputs)
  
}

```

### Test GLMM function

```{r}
response_variable <- "sprich"
model_structure <- "~ treatment*year + (1|block/plot)"
dataset <- analysis_df

fit_and_check_glmm_model(
  response_variable = response_variable,
  model_structure = model_structure,
  dataset = analysis_df
)

```

## Fit and check LMM models function

```{r}
fit_and_check_lmm_model <- function(response_variable, model_structure, dataset) {
  
  # Create list to store all outputs for one model
  model_outputs <<- list()
  
  model_outputs[["response_variable"]] <<- response_variable

  # Create the formula for the linear mixed effects model
  formula <- as.formula(paste(response_variable, model_structure))
  print(formula)
  
  model_outputs[["formula"]] <<- formula

  # Fit the linear mixed effects model and store it in a list
  fitted_model <- lmer(formula, data = dataset)
    
  model_outputs[["fitted_model"]] <<- fitted_model
  
  print(summary(fitted_model))
  
  # Validate model
  check_my_model(fitted_model, dataset = dataset)
  
  # Other diagnostics
  print(check_model(fitted_model))
  
  # Test
  test_my_model(fitted_model)
  
  # Simulate
  simulate_my_model(fitted_model, rv = response_variable)
  
  return(model_outputs)
}
```

### Test LMM function.

```{r}
response_variable <- "shannon"
model_structure <- "~ treatment*year + (1|block/plot)"
dataset = analysis_df

fit_and_check_lmm_model(
  response_variable = response_variable,
  model_structure = model_structure,
  dataset = analysis_df
)

```

## Run models individually to assess fit and diagnostics

### Species richness

Try model structures with and without block (to account for spatial autocorrelation).

#### Year as categorical

```{r}
response_variable <- "sprich"
model_structure <- c("~ treatment*year + (1|block/plot)", "~ treatment*year + (1|plot)")
dataset = analysis_df

for (structure in model_structure) {
  
  print("_______________________________________________________")
  
  fit_and_check_glmm_model(
    response_variable = response_variable,
    model_structure = structure,
    dataset = analysis_df
  )
  
  print("_______________________________________________________")
}

```

#### Year as continuous

```{r}
sprich_m1 <- fit_and_check_glmm_model(response_variable = "sprich",
    model_structure = "~ treatment*year_num + (1|block/plot)",
    dataset = analysis_df)

sprich_m1[['fitted_model']]
test_my_model(sprich_m1[['fitted_model']])

sjPlot::plot_model(sprich_m1[['fitted_model']], type = "pred", terms = c("year_num", "treatment"))
```

Model using negative binomial distribution (Cullen-Frey graph indicates this distribution would also work).

```{r}
sprich_m2 <- glmmTMB(sprich ~ treatment*year + (1|block/plot),
                        data = analysis_df,
                        family= nbinom2
)

summary(sprich_m2)
simulate_my_model(sprich_m2, rv = "sprich")

test_my_model(sprich_m2)
```

### Native species richness

#### Year as categorical

```{r}
response_variable <- "sprich_native"
model_structure <- c("~ treatment*year + (1|block/plot)", "~ treatment*year + (1|plot)")
dataset = analysis_df

for (structure in model_structure) {
  
  print("_______________________________________________________")
  
  fit_and_check_glmm_model(
    response_variable = response_variable,
    model_structure = structure,
    dataset = analysis_df
  )
  
  print("_______________________________________________________")
}
```

#### Year as continuous

```{r}
sprich_native_m1 <- fit_and_check_glmm_model(response_variable = "sprich_native",
    model_structure = "~ treatment*year_num + (1|block/plot)",
    dataset = analysis_df)

sprich_native_m1[['fitted_model']]
test_my_model(sprich_native_m1[['fitted_model']])

sjPlot::plot_model(sprich_native_m1[['fitted_model']], type = "pred", terms = c("year_num", "treatment"))
```

### Shannon

#### Year as categorical

```{r}
response_variable <- "shannon"
model_structure <- c("~ treatment*year + (1|block/plot)", "~ treatment*year + (1|plot)")
dataset = analysis_df

for (structure in model_structure) {
  
  print("_______________________________________________________")
  
  fit_and_check_lmm_model(
    response_variable = response_variable,
    model_structure = structure,
    dataset = analysis_df
  )
  
  print("_______________________________________________________")
}
```

```{r}
sh_m1 <- fit_and_check_lmm_model(
    response_variable = "shannon",
    model_structure = "~ treatment*year + (1|block/plot)",
    dataset = analysis_df
    )

simulate_my_model(sh_m1[['fitted_model']], rv = "shannon")

test_my_model(sh_m1[['fitted_model']])

sjPlot::plot_model(sh_m1[['fitted_model']], type = "pred", terms = c("year", "treatment"))

```

#### Year as continuous

```{r}
sh_m2 <- fit_and_check_lmm_model(response_variable = "shannon",
    model_structure = "~ treatment*year_num + (1|block/plot)",
    dataset = analysis_df)

sh_m2[['fitted_model']]
test_my_model(sh_m2[['fitted_model']])

sjPlot::plot_model(sh_m2[['fitted_model']], type = "pred", terms = c("year_num", "treatment"))

simulate_my_model(sh_m2[['fitted_model']], rv = "shannon")
```

### Shannon native

#### Year as categorical

```{r}
response_variable <- "shannon_native"
model_structure <- c("~ treatment*year + (1|block/plot)" 
                     # "~ treatment*year + (1|plot)"
                     )
dataset = analysis_df

for (structure in model_structure) {
  
  print("_______________________________________________________")
  
  fit_and_check_lmm_model(
    response_variable = response_variable,
    model_structure = structure,
    dataset = analysis_df
  )
  
  print("_______________________________________________________")
}
```

This data is clearly zero-inflated.

#### Log transform

What if we log transform?

```{r}
response_variable <- "log_shannon_native"
model_structure <- "~ treatment*year + (1|block/plot)"
dataset = analysis_df


fit_and_check_lmm_model(
  response_variable = response_variable,
  model_structure = model_structure,
  dataset = analysis_df
)

```

Still problematic diagnostics.

#### ZIF, year as categorical

What if we try a zero-inflated model?

We will use a beta distribution as the Cullen-Frey graph indicates this is the best fit for the data (see 3_data_exploration).

```{r}
analysis_df <- analysis_df %>%
  mutate(norm_shannon_native = ((((shannon_native - min(shannon_native)) 
                                 / max(shannon_native) - min(shannon_native))/1.01))
         )
```

Model with zero inflation.

```{r}
shnat_m1 <- glmmTMB(log_shannon_native ~ treatment*year + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family = beta_family())

simulate_my_model(model = shnat_m1, rv = "shannon_native")

check_predictions(shnat_m1)

summary(shnat_m1)

check_my_model(shnat_m1, dataset = analysis_df)

check_model(shnat_m1)

plot(simulateResiduals(fittedModel = shnat_m1))

sjPlot::plot_model(shnat_m1, type = "pred", terms = c("treatment", "year"), show.values = TRUE)

car::Anova(shnat_m1, component = "cond")
```

#### ZIF, year as continuous

Model with year as continuous.

```{r}
shnat_m2 <- glmmTMB(log_shannon_native ~ treatment*year_num + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family = beta_family())

simulate_my_model(model = shnat_m2, rv = "shannon_native")

check_predictions(shnat_m2)

summary(shnat_m2)

check_model(shnat_m2)

plot(simulateResiduals(fittedModel = shnat_m2))


sjPlot::plot_model(shnat_m2, type = "pred", terms = c("year_num", "treatment"), show.values = TRUE)

car::Anova(shnat_m2, component = "cond")

```

### Exotic forb

#### Year as categorical

```{r}
ef_m1 <- glmmTMB(exotic_forb ~ treatment*year + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family= beta_family()
)

summary(ef_m1)

check_model(ef_m1)

plot(simulateResiduals(fittedModel = ef_m1))

simulate_my_model(ef_m1, rv = "exotic_forb")

test_my_model(ef_m1)

sjPlot::plot_model(ef_m1, type = "pred", terms = c("year", "treatment"), show.values = TRUE)

car::Anova(ef_m1, component = "cond")

```

#### Year as continuous

```{r}
ef_m2 <- glmmTMB(exotic_forb ~ treatment*year_num + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family= beta_family()
)

summary(ef_m2)

check_model(ef_m2)
simulate_my_model(model = ef_m2, rv = "exotic_forb")

test_my_model(ef_m2)

sjPlot::plot_model(ef_m2, type = "pred", terms = c("year_num", "treatment"),
                   show.values = TRUE)

car::Anova(ef_m2, component = "cond")

```

### Native forb

#### Year as categorical

```{r}
nf_m1 <- glmmTMB(native_forb ~ treatment*year + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family= beta_family()
)

summary(nf_m1 )

check_model(nf_m1 )

plot(simulateResiduals(fittedModel = nf_m1 ))

simulate_my_model(nf_m1, rv = "native_forb")

sjPlot::plot_model(nf_m1 , type = "pred", terms = c("year", "treatment"), show.values = TRUE)

car::Anova(nf_m1, component = "cond")

```

#### Year as continuous

```{r}
nf_m2 <- glmmTMB(native_forb ~ treatment*year_num + (1|block/plot),
                        ziformula=~1,
                        data = analysis_df,
                        family= beta_family()
)

summary(nf_m2)

check_model(nf_m2)

plot(simulateResiduals(fittedModel = nf_m2))

simulate_my_model(nf_m2, rv = "native_forb")

sjPlot::plot_model(nf_m2, type = "pred", terms = c("year_num", "treatment"), show.values = TRUE)

car::Anova(nf_m2, component = "cond")
```

Compare to a regular linear mixed model.

```{r}
nf_lmm <- fit_and_check_lmm_model(
  response_variable = "native_forb",
  model_structure = "~ treatment*year + (1|block/plot)",
  dataset = analysis_df
)

test_my_model(nf_lmm[['fitted_model']])
```

