---
title: "Single Year Models"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot)
library(lme4)
library(glmmTMB)
library(car)
library(emmeans)
library(stringr)
library(performance)
library(purrr)
library(gridExtra)

# options(scipen = 999)
```


```{r}
analysis_df <- read.csv("analyses_manuscript/processed_data/analysis_variables.csv") %>%
  mutate(year = as.factor(year)) %>%
  mutate(log_shannon_native = log(shannon_native + 1)) %>%
  mutate(shannon_native_backtransform = exp(log_shannon_native)-1)

analysis_2022_df <- analysis_df %>%
  filter(year == "2022")
  
analysis_2023_df <- analysis_df %>%
  filter(year == "2023")
```

## Function to test models

First, a function to test the model and return the test itself.

```{r}
test_single_year_model <- function(fitted_model) {
  
  model_results <- list()
  
  chisqAnova <- car::Anova(fitted_model)
  print(chisqAnova)
  model_results[["chisquare_test"]] <- chisqAnova 
  
  # Extract p-value for the treatment*year interaction term
  pval <- chisqAnova$'Pr(>Chisq)'[1]
  pval.r <- ifelse(
    pval < 0.001,
    "< 0.001",
    round(pval, digits = 3)
  )

  # Assess significance of p-value
  significance <- ifelse(pval < 0.05,
                       "*",
                       "N.S.")
  pval_signif <- paste(pval.r, significance, sep = " ")
  
  print(pval_signif)
  model_results[["significance"]] <- pval_signif
  
  return(model_results)
}
```

Now, a function to save the results to the global environment (every time this function is called, it will also save the results to a list in the global environment).

```{r}
# Initialize a list to store all model results
single_year_results_list <- list()

save_model_results <- function(fitted_model, year) {
  
  model_results <- test_single_year_model(fitted_model = fitted_model)
  
  # Every time this function is called, it will add the results to the list of 
  # results in the global environment
  model_name <- deparse(substitute(fitted_model))
  model_name_with_year <- paste(model_name, year, sep = "_")
  single_year_results_list[[model_name_with_year]] <<- model_results
  
}

```


## Function to plot models

```{r}
plot_single_year_model <- function(fitted_model, year, response_variable, response_name, dataset) {

  # Test the model
  model_results <- test_single_year_model(fitted_model = fitted_model)
  
  significance <- model_results[['significance']]
  
  # Calculate position for text label
  # rv_vec <- analysis_df[[response_variable]]
  # label_position_x <- 'unraked'
  # label_position_y <- max(rv_vec) + (max(rv_vec) * 0.05)
  
  # Format label
  significance_label <- paste("p-value:", significance, sep = " ")
  
  # Plot title
  # plot_title <- paste(year, sep = " ")
  plot_title <- significance_label
    
  # Make plot
  model_plot <- dataset %>%
    ggplot(aes_string(
      x = "treatment",
      y = response_variable,
      fill = "treatment",
      color = "treatment",
      group = "treatment"
    )) +
    geom_jitter(
                alpha = 0.3,
                shape = 16,
                width = 0.3,
                height = 0,
                size = 3
                ) +
    geom_boxplot(color = "black", alpha = 0.4, outlier.shape = NA) +
    scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
    scale_color_manual(
      values = c("antiquewhite4", "darkgoldenrod2"),
      guide = guide_legend(override.aes = list(alpha = 1))
    ) +
  labs(title = plot_title) +
  theme_pubr() +
  theme(legend.position = "none",
        plot.title = element_text(size = 12)) +
  ylab(response_name)
  # geom_text(aes(x = label_position_x,
  #               y = label_position_y),
  # label = significance_label,
  # hjust = 1,
  # vjust = 1,
  # size = 3,
  # color = "black")

  return(model_plot)
}

```

# Individual year models

### Species richness

```{r}
fit_sprich <- function(dataset, year) {
  
  sprich_fm <- glmer(sprich ~ treatment + (1 | block),
                     family = "poisson",
                     data = dataset)
  
  save_model_results(sprich_fm, year)
  
  p <- plot_single_year_model(
    fitted_model = sprich_fm,
    year = year,
    response_variable = "sprich",
    response_name = "Species richness",
    dataset = dataset
  )
  
  return(p)
}

```


```{r}
p_sprich_2022 <- fit_sprich(dataset = analysis_2022_df, year = "2022")
p_sprich_2023 <- fit_sprich(dataset = analysis_2023_df, year = "2023")
```

### Native species richness

```{r}
fit_sprich_native <- function(dataset, year) {
  
  sprich_native_fm <- glmer(sprich_native ~ treatment + (1 | block),
                     family = "poisson",
                     data = dataset)
  
  save_model_results(sprich_native_fm, year)
  
  p <- plot_single_year_model(
    fitted_model = sprich_native_fm,
    year = year,
    response_variable = "sprich_native",
    response_name = "Native species richness",
    dataset = dataset
  )
  
  return(p)
}

```

```{r}
p_sprich_native_2022 <- fit_sprich_native(dataset = analysis_2022_df, year = "2022")
p_sprich_native_2023 <- fit_sprich_native(dataset = analysis_2023_df, year = "2023")
```


```{r}
fit_shannon <- function(dataset, year) {
  
  shannon_fm <- lmer(shannon ~ treatment + (1|block),
                   data = dataset)
  
  save_model_results(shannon_fm, year)
  
  p <- plot_single_year_model(
    fitted_model = shannon_fm,
    year = year,
    response_variable = "shannon",
    response_name = "Shannon diversity",
    dataset = dataset
  )

  return(p)
}
```

```{r}
p_shannon_2022 <- fit_shannon(dataset = analysis_2022_df, year = "2022")
p_shannon_2023 <- fit_shannon(dataset = analysis_2023_df, year = "2023")
```

### Shannon Native

```{r}
fit_shannon_native <- function(dataset, year) {
  
  shnat_fm <- glmmTMB(log_shannon_native ~ treatment + (1|block),
                        ziformula=~1,
                        data = dataset,
                        family = beta_family())
  
  save_model_results(shnat_fm, year)
  
  p <- plot_single_year_model(
    fitted_model = shnat_fm,
    year = year,
    response_variable = "shannon_native",
    response_name = "Native Shannon diversity",
    dataset = dataset
  )

  return(p)
}
```

```{r}
p_shannon_native_2022 <- fit_shannon_native(dataset = analysis_2022_df, year = "2022")
p_shannon_native_2023 <- fit_shannon_native(dataset = analysis_2023_df, year = "2023")
```

### Exotic forb

```{r}
fit_exotic_forb <- function(dataset, year) {
  
  ef_fm <- glmmTMB(exotic_forb ~ treatment + (1|block),
                        ziformula=~1,
                        data = dataset,
                        family= beta_family()
                   )
  
  save_model_results(ef_fm, year)
  
  p <- plot_single_year_model(
    fitted_model = ef_fm,
    year = year,
    response_variable = "exotic_forb",
    response_name = "Exotic forb abundance",
    dataset = dataset
  )

  return(p)
}
```

```{r}
p_exotic_forb_2022 <- fit_exotic_forb(dataset = analysis_2022_df, year = "2022")
p_exotic_forb_2023 <- fit_exotic_forb(dataset = analysis_2023_df, year = "2023")
```

### Native forb

```{r}
fit_native_forb <- function(dataset, year) {
  
  nf_fm <- glmmTMB(native_forb ~ treatment + (1|block),
                        ziformula=~1,
                        data = dataset,
                        family= beta_family()
)
  
  save_model_results(nf_fm, year)
  
  p <- plot_single_year_model(
    fitted_model = nf_fm,
    year = year,
    response_variable = "native_forb",
    response_name = "Native forb abundance",
    dataset = dataset
  )

  return(p)
}
```

```{r}
p_native_forb_2022 <- fit_native_forb(dataset = analysis_2022_df, year = "2022")
p_native_forb_2023 <- fit_native_forb(dataset = analysis_2023_df, year = "2023")
```

## Format and export plots

```{r}
arrange_plots <- function(title, plots_list, ymin, ymax) {
  
  plots_adjusted_list <- list()

    for (plot_name in names(plots_list)) {
    
    plot <- plots_list[[plot_name]]
    plot_adj <- plot + ylim(min = ymin, max = ymax)
    
    plots_adjusted_list[[plot_name]] <- plot_adj
}
  return(plots_adjusted_list)
  
}
```


```{r}
sprich_as_unit_plots <- list(
  p_sprich_2022 = p_sprich_2022,
  p_sprich_2023 = p_sprich_2023,
  p_sprich_native_2022 = p_sprich_native_2022,
  p_sprich_native_2023= p_sprich_native_2023)

sprich_as_unit_plots_adj <- arrange_plots(
  title = "Species richness",
  plots_list = sprich_as_unit_plots, ymin = 0, ymax = 14)

sprich_as_unit_plots_adj

shannon_as_unit_plots <- list(
  p_shannon_2022 = p_shannon_2022,
  p_shannon_2023= p_shannon_2023,
  p_shannon_native_2022 = p_shannon_native_2022,
  p_shannon_native_2023 =  p_shannon_native_2023)

shannon_as_unit_plots_adj <- arrange_plots(
  title = "Shannon diversity",
  plots_list = shannon_as_unit_plots, ymin = 0, ymax = 2)

shannon_as_unit_plots_adj

abundance_as_unit_plots <- list(
  p_exotic_forb_2022 = p_exotic_forb_2022,
  p_exotic_forb_2023 = p_exotic_forb_2023,
  p_native_forb_2022 = p_native_forb_2022,
  p_native_forb_2023 = p_native_forb_2023)

abundance_as_unit_plots_adj <- arrange_plots(
  title = "Abundance",
  plots_list = abundance_as_unit_plots, ymin = 0, ymax = 1)

abundance_as_unit_plots_adj
```

```{r}
plots_by_unit_list <- list(sprich_as_unit_plots_adj,
shannon_as_unit_plots_adj,
abundance_as_unit_plots_adj)

all_plots_list <- purrr::flatten(plots_by_unit_list)

names(all_plots_list)


```

```{r}

order_plots_for_2022 <- c(
  "p_sprich_2022", "p_sprich_native_2022", "p_shannon_2022", "p_shannon_native_2022", "p_exotic_forb_2022", "p_native_forb_2022"
)

order_plots_for_2023 <- c(
  "p_sprich_2023", "p_sprich_native_2023", "p_shannon_2023", "p_shannon_native_2023", "p_exotic_forb_2023", "p_native_forb_2023"
)

plots_for_2022 <- all_plots_list[order_plots_for_2022]

plots_for_2023 <- all_plots_list[order_plots_for_2023]
```


```{r}
plots_combined_2022 <- grid.arrange(grobs = plots_for_2022, ncol = 6, nrow = 1, left = "2022")

plots_combined_2023 <- grid.arrange(grobs = plots_for_2023, ncol = 6, nrow = 1, left = "2023")

arranged_single_year_model_plots <- grid.arrange(plots_combined_2022, plots_combined_2023)
arranged_single_year_model_plots
```

```{r}
ggsave(filename = "arranged_single_year_model_plots.pdf",
         plot = arranged_single_year_model_plots,
         path = "analyses_manuscript/figures",
         width = 14,
         height = 5,
          scale = 1.2)
```

Save the results as a table.

```{r}
# Save table of significance.

all_results_df <- data.frame(
  model = character(),
  predictor = character(),
  Chisq = numeric(),
  Df = numeric(), 
  p_value = numeric()
)

for (model_name in names(single_year_results_list)) {
  
  anova_table <- single_year_results_list[[model_name]][['chisquare_test']]
  
  anova_table_formatted <- anova_table %>%
    rownames_to_column("predictor") %>%
    rename("p_value" = `Pr(>Chisq)`) %>%
    mutate(model = model_name) %>%
    mutate(p_value_f = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value > 0.001 ~ paste(
        signif(p_value, digits = 2))
    )) %>%
    mutate(significance = case_when(
      p_value < 0.05 ~ "*",
      p_value > 0.05 ~ "N.S."
    ))
  
  all_results_df <- bind_rows(all_results_df, anova_table_formatted)

}

all_results_df
  
```

```{r}
write.csv(
  all_results_df,
  file = "analyses_manuscript/stats_results/single_year_model_results.csv",
  row.names = FALSE
)
```

