---
title: "Cumulative raking models"
author: "Advyth Ramachandran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: journal
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Note**

The models in this analysis were not used to draw inferences. This was due to the limited information provided by using years raked as a continuous variable, which conflates differences among years in annual precipitation with differences due to possible cumulative effects of raking (i.e., compounding effects of multiple years of litter removal).

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot)
library(ggeffects)
library(lme4)
library(glmmTMB)
library(car)
library(emmeans)
library(stringr)
library(performance)
library(purrr)
library(gridExtra)

load("analyses_manuscript/theme_function.R")
```


```{r}
analysis.df <- read.csv("analyses_manuscript/processed_data/analysis_variables.csv") %>%
  mutate(year = as.factor(year)) %>%
  mutate(log_shannon_native = log(shannon_native + 1)) %>%
  mutate(shannon_native_backtransform = exp(log_shannon_native)-1)

analysis.only.raked.df <- analysis.df %>%
  filter(treatment == "raked")
```

## Function to test models

```{r}
test_cumulative_raking_model <- function(fitted_model) {
  
  chisqAnova <- car::Anova(fitted_model)
  
  print(chisqAnova)
  
  # Extract p-value for the treatment*year interaction term
  pval <- chisqAnova$'Pr(>Chisq)'[1]
  pval.r <- ifelse(
    pval < 0.001,
    "< 0.001",
    round(pval, digits = 3)
  )

  # Assess significance of p-value
  significance <- ifelse(pval < 0.05,
                       "*",
                       "N.S.")
  pval_signif <- paste(pval.r, significance, sep = " ")
    
  print(pval_signif)
  
  return(pval_signif)
}
```

## Function to plot models

```{r}
plot_cumulative_raking_model <- function(fitted_model, response_variable, response_name, dataset, pval) {
  
  # Check validity
  print(check_model(fitted_model))
  
  # Calculate p-value
  pval = test_cumulative_raking_model(fitted_model)

  # Format p-value label
  significance_label <- paste("p-value:", pval, sep = " ")
  # rv_vec <- dataset[[response_variable]]
  # label_position_x <- 2
  # label_position_y <- max(rv_vec) + (max(rv_vec) * 0.05)
  
  # Format plot title
  plot_title <- paste(
    paste(response_name, ",", sep = ""),
          significance_label, sep = " ") # if a different format is desired
  # plot_title <- significance_label
  
  # Compute predicted values
  predicted_values <- ggpredict(fitted_model, terms = "years_raked")

  # Make plot
  model_plot <- plot(predicted_values) +
    geom_jitter(
      data = dataset,
      aes_string(
        x = "years_raked",
        y = response_variable,
        fill = "treatment",
        color = "treatment"
        ),
      alpha = 0.3,
      shape = 16,
      width = 0.1,
      height = 0,
      size = 3
      ) +
    scale_fill_manual(values = c("antiquewhite4", "darkgoldenrod2")) +
    scale_color_manual(
      values = c("antiquewhite4", "darkgoldenrod2"),
      guide = guide_legend(override.aes = list(alpha = 1))
    ) +
  # labs(title = plot_title) +
  theme_Publication() +
  theme(legend.position = "none",
        plot.title = element_text(size = 12)) +
  ylab(response_name) +
  xlab("years of raking") +
  labs(title = NULL) +
  scale_x_continuous(breaks = c(0, 1, 2)) +
  # If label should be added as an inset:
  # geom_text(aes(x = label_position_x,
  #               y = label_position_y),
  # label = significance_label,
  # hjust = 1,
  # vjust = 1,
  # size = 4,
  # color = "black") +
  annotate("text", x = Inf, y = Inf, label = paste(significance_label), vjust = 2, hjust = 2)


  return(model_plot)
}

```


# Individual year models

### Species richness

```{r}

sprich_fm <- glmer(sprich ~ years_raked + (1|plot),
                     family = "poisson",
                     data = analysis.only.raked.df)
  
print(sprich_fm)

p_sprich_cml <- plot_cumulative_raking_model(
  fitted_model = sprich_fm,
  response_variable = "sprich",
  response_name = "Species richness",
  dataset = analysis.only.raked.df
)
p_sprich_cml
```

### Native species richness

```{r}
  
sprich_native_fm <- glmer(sprich_native ~ years_raked + (1|plot),
                   family = "poisson",
                   data = analysis.only.raked.df)
  
p_sprich_native_cml <- plot_cumulative_raking_model(
  fitted_model = sprich_native_fm,
  response_variable = "sprich_native",
  response_name = "Native species richness",
  dataset = analysis.only.raked.df,
)
p_sprich_native_cml

```

```{r}

shannon_fm <- lmer(shannon ~ years_raked + (1|plot),
                 data = analysis.only.raked.df)

p_shannon_cml <- plot_cumulative_raking_model(
  fitted_model = shannon_fm,
  response_variable = "shannon",
  response_name = "Shannon diversity",
  dataset = analysis.only.raked.df
)
p_shannon_cml
```

### Shannon Native

```{r}

shnat_fm <- glmmTMB(log_shannon_native ~ years_raked + (1|plot),
                      ziformula=~1,
                      data = analysis.only.raked.df,
                      family = beta_family())
  
p_shannon_native_cml <- plot_cumulative_raking_model(
  fitted_model = shnat_fm,
  response_variable = "shannon_native",
  response_name = "Native Shannon diversity",
  dataset = analysis.only.raked.df
)
p_shannon_native_cml
```

### Exotic forb

```{r}

ef_fm <- glmmTMB(exotic_forb ~ years_raked + (1|plot),
                      ziformula=~1,
                      data = analysis.only.raked.df,
                      family= beta_family()
                 )

p_exotic_forb_cml <- plot_cumulative_raking_model(
  fitted_model = ef_fm,
  response_variable = "exotic_forb",
  response_name = "Exotic forb abundance",
  dataset = analysis.only.raked.df
)
p_exotic_forb_cml

```

### Native forb

```{r}
  
nf_fm <- glmmTMB(native_forb ~ years_raked + (1|plot),
                      ziformula=~1,
                      data = analysis.only.raked.df,
                      family= beta_family()
)

p_native_forb_cml <- plot_cumulative_raking_model(
  fitted_model = nf_fm,
  response_variable = "native_forb",
  response_name = "Native forb abundance",
  dataset = analysis.only.raked.df
)

p_native_forb_cml

```

## Format and export plots

```{r}
arrange_plots <- function(title, plots_list, ymin, ymax) {
  
  plots_adjusted_list <- list()

    for (plot_name in names(plots_list)) {
    
    plot <- plots_list[[plot_name]]
    plot_adj <- plot + ylim(min = ymin, max = ymax)
    
    plots_adjusted_list[[plot_name]] <- plot_adj
}
  return(plots_adjusted_list)
  
}
```


```{r}
# Put all the variables with the same y-axis unit on the same y-scale

sprich_as_unit_plots <- list(
  p_sprich_cml = p_sprich_cml,
  p_sprich_native_cml = p_sprich_native_cml)

sprich_as_unit_plots_adj <- arrange_plots(
  plots_list = sprich_as_unit_plots, ymin = 0, ymax = 14)

sprich_as_unit_plots_adj

shannon_as_unit_plots <- list(
  p_shannon_cml = p_shannon_cml,
  p_shannon_native_cml = p_shannon_native_cml)

shannon_as_unit_plots_adj <- arrange_plots(
  plots_list = shannon_as_unit_plots, ymin = 0, ymax = 2)

shannon_as_unit_plots_adj

abundance_as_unit_plots <- list(
  p_exotic_forb_cml = p_exotic_forb_cml,
  p_native_forb_cml = p_native_forb_cml)

abundance_as_unit_plots_adj <- arrange_plots(
  plots_list = abundance_as_unit_plots, ymin = 0, ymax = 1)

abundance_as_unit_plots_adj

# Combine all plots into a single list
plots_by_unit_list <- list(sprich_as_unit_plots_adj,
shannon_as_unit_plots_adj,
abundance_as_unit_plots_adj)

all_plots_list <- purrr::flatten(plots_by_unit_list)

names(all_plots_list)
```

```{r}
# Define plot order to be displayed

order_of_plots <- c(
  "p_sprich_cml", # not significant in global
  "p_sprich_native_cml", # not significant in global
  "p_shannon_cml",
  "p_shannon_native_cml", # not significant in global
  "p_exotic_forb_cml",
  "p_native_forb_cml" # not significant in global
)

ordered_plots <- all_plots_list[order_of_plots]

```

```{r}
# Arrange plots
arranged_cumulative_raking_plots <- grid.arrange(grobs = ordered_plots, ncol = 2, nrow = 3)
arranged_cumulative_raking_plots

```

```{r}
ggsave(filename = "arranged_cumulative_raking_model_plots.pdf",
         plot = arranged_cumulative_raking_plots,
         path = "analyses_manuscript/figures",
         width = 8,
         height = 8,
          scale = 1.2)
```

```{r}
order_of_inference_plots <- c(
  # "p_sprich_cml", # not significant in global
  # "p_sprich_native_cml", # not significant in global
  "p_shannon_cml", 
  # "p_shannon_native_cml", # not significant in global
  "p_exotic_forb_cml"
  # "p_native_forb_cml" # not significant in global
)

ordered_inference_plots <- all_plots_list[order_of_inference_plots]

# Arrange plots
arranged_inference_cumulative_raking_plots <- grid.arrange(grobs = ordered_inference_plots, ncol = 2, nrow = 1)
arranged_inference_cumulative_raking_plots
```

